---
title: "Leave-One-Out Brier Score for survival models"
subtitle: "Using Pareto smoothed importance sampling"
author: "Eren M. El&ccedil;i"
date: "2018-11-21"
output:
  tufte::tufte_html: default
  tufte::tufte_handout: default

---

In this document we show how to implement a leave-one-out version of the so-called "inverse of probability of censoring weighted estimator" (below referred to as Brier Score) for time dependent prediction accuracy of Bayesian survival models.

We use Pareto-Smoothed importance sampling here in order to avoid re-running the Bayesian analysis for any leave-one-out instance.

We provide an efficient implementation written in Stan and exposed to R.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(rstan);rstan_options(auto_write = TRUE)
library(loo)
library(survival)
library(tibble)
library(dplyr)
library(purrr)
library(bayesplot)
library(splines2)
library(pec)
library(HSAUR)
library(cowplot)
cbbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
expose_stan_functions("~/Desktop/Stan/Loo_Pec/utils_surv.stan")
sm <- stan_model("~/Desktop/Stan/Loo_Pec/exponential_survival_simple_ppc_cond_loo.stan")
sm2 <- stan_model("~/Desktop/Stan/Loo_Pec/survival_parametric_baseline_hazard_simplex_loo.stan")
data("mastectomy")
df <- as.tibble(mastectomy)
df <- df %>% mutate(metastized=as.double(metastized=="yes"))
```

```{r inspect, echo=FALSE}
knitr::kable(
  df, caption = 'A mastectomy dataset'
)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE, error=FALSE}
N <- nrow(df)
X <- as.matrix(as.integer(pull(df, metastized))-1)
is_censored <- pull(df,event)==0
times <- pull(df,time)
msk_censored <- is_censored == 1
N_censored <- sum(msk_censored)

time_range <- range(times)
time_min <- time_range[1]
time_max <- time_range[2] #time_range[2]

knots <- quantile(times[!msk_censored], probs = c(.05, .35, .65, .95))

nknots <- length(knots)
mspline_degree<- 3
i_spline_basis_evals <- iSpline(times, knots=knots, degree=mspline_degree,
                                intercept=FALSE, Boundary.knots = c(0, time_max))
m_spline_basis_evals <- deriv(i_spline_basis_evals)
i_spline_basis_evals_censored <- i_spline_basis_evals[msk_censored,]
i_spline_basis_evals_uncensored <- i_spline_basis_evals[!msk_censored,]
m_spline_basis_evals_uncensored <- m_spline_basis_evals[!msk_censored,]
nbasis <- dim(i_spline_basis_evals_censored)[2]




stan_data <- list(N_uncensored=N-N_censored, 
                  N_censored=N_censored, 
                  X_censored=as.matrix(X[msk_censored,]),
                  X_uncensored=as.matrix(X[!msk_censored,]),
                  times_censored=times[msk_censored],
                  times_uncensored = times[!msk_censored],
                  NC=ncol(X),
                  m=nbasis,
                  m_spline_basis_evals_uncensored=m_spline_basis_evals_uncensored, 
                  i_spline_basis_evals_uncensored=i_spline_basis_evals_uncensored,
                  i_spline_basis_evals_censored=i_spline_basis_evals_censored
)

fit <- sampling(sm, data=stan_data, seed=42, chains=4, cores=2, iter=4000)
fit2 <- sampling(sm2, data=stan_data, seed=42, chains=4, cores=2, iter=4000)

post <- as.array(fit)
post2 <- as.array(fit2)
```


## Loo

```{r loo, echo=TRUE}
log_lik_1 <- extract_log_lik(fit, merge_chains = FALSE)
r_eff <- relative_eff(exp(log_lik_1)) 
loo_1 <- loo(log_lik_1, r_eff = r_eff, cores = 2)
print(loo_1)

log_lik_2 <- extract_log_lik(fit2, merge_chains = FALSE)
r_eff <- relative_eff(exp(log_lik_2)) 
loo_2 <- loo(log_lik_2, r_eff = r_eff, cores = 2)
print(loo_2)
comp <- compare(loo_1, loo_2)

print(comp)
```

## Survival curves

## M-Splines

```{r surv, echo=FALSE}
intercept = summary(fit2)$summary["intercept", "mean"]
betas = summary(fit2)$summary["betas[1]", "mean"]
gammas = summary(fit2)$summary[sprintf("gammas[%d]", 1:nbasis),"mean"]
X_surv = as.matrix(c(0,1))

times_plot <- seq(0, time_max, length.out = 100)
i_spline_basis_evals <- iSpline(times_plot, knots=knots, degree=mspline_degree,
                                intercept=FALSE, Boundary.knots = c(0, time_max))

survs <- surv_mspline(i_spline_basis_evals, X_surv, gammas, intercept, betas)

surv_df <- as.tibble(t(survs)) %>%
  mutate(t=times_plot) %>%
  rename(x0=V1, x1=V2)

surv_df <- bind_rows(dplyr::select(surv_df, t, x0) %>% rename(surv=x0) %>% mutate(metastized=FALSE), 
                     dplyr::select(surv_df, t, x1)  %>% rename(surv=x1) %>% mutate(metastized=TRUE))
knitr::kable(
  head(surv_df,10)
)

ggplot(data=surv_df) + 
  geom_vline(xintercept = c(0,knots, time_max), alpha=.5, color='gray', linetype='dashed')+
  geom_line(mapping=aes(x=t, y=surv, color=metastized))+
  geom_rug(data=tibble(times=times), aes(x=times), col="steelblue",alpha=0.5, size=1.25)+
  scale_fill_manual(values=cbbPalette)+
  scale_colour_manual(values=cbbPalette)
```

## Constant

```{r surv_constant, echo=FALSE}
intercept = summary(fit)$summary["intercept", "mean"]
betas = summary(fit)$summary["betas[1]", "mean"]
X_surv = as.matrix(c(0,1))

times_plot <- seq(0, time_max, length.out = 100)


survs <- surv_const(times_plot, X_surv, intercept, betas)

surv_df <- as.tibble(t(survs)) %>%
  mutate(t=times_plot) %>%
  rename(x0=V1, x1=V2)

surv_df <- bind_rows(dplyr::select(surv_df, t, x0) %>% rename(surv=x0) %>% mutate(metastized=FALSE), 
                     dplyr::select(surv_df, t, x1)  %>% rename(surv=x1) %>% mutate(metastized=TRUE))
knitr::kable(
  head(surv_df,10)
)

ggplot(data=surv_df) + 
  geom_vline(xintercept = c(0,knots, time_max), alpha=.5, color='gray', linetype='dashed')+
  geom_line(mapping=aes(x=t, y=surv, color=metastized))+
  geom_rug(data=tibble(times=times), aes(x=times), col="steelblue",alpha=0.5, size=1.25)+
  scale_fill_manual(values=cbbPalette)+
  scale_colour_manual(values=cbbPalette)
```

## Brier Score (LOO-PSIS)

$$
S_i(t\vert \mathbf{T}_{-i}) \approx \frac{\sum_{s=1}^S r_{s,i} S_i(t\vert \theta_s)}{\sum_{s=1}^S r_{s,i}}
$$

with

$$
r_{s,i} \equiv \frac{1}{p(T_i\vert \theta_s)}
$$

$$
\eta_{i,t} \equiv \left\{ 1_{\{T_i > t\}} - S_i(t\vert \mathbf{T}_{-i})\right\}^2
$$

Based on the residuals $\eta_{i,t}$ using the `pec` R Package, we calculate the so-called inverse of probability of censoring weighted estimator (below referred to as Brier Score)

```{r, echo=FALSE}
times_pec <- seq(0, time_max, length.out = 100)
log_ratios <- -post[,,sprintf("log_lik[%d]",1:N)]
r_eff <- relative_eff(exp(log_ratios))
psis_result <- psis(log_ratios, r_eff = r_eff)
w <- weights(psis_result, log=FALSE) #note that colSums(w2) = rep(1, dim(w2)[2])
survs_loo1 <- surv_const_loo(times_pec,
                            X,
                            as.vector(post[,,"intercept"]),
                            matrix(post[,,sprintf("betas[%d]" ,1:ncol(X))], dim(post)[1]*dim(post)[2], ncol(X)),
                            t(w)
                            )
survs1 <- surv_const_loo(times_pec,
                            X,
                            as.vector(post[,,"intercept"]),
                            matrix(post[,,sprintf("betas[%d]" ,1:ncol(X))], dim(post)[1]*dim(post)[2], ncol(X)),
                            matrix(rep(1, dim(w)[1]*dim(w)[2]), nrow = dim(w)[2], byrow = T)/dim(w)[1]
                            )
```

```{r, echo=FALSE}
i_spline_basis_evals <- iSpline(times_pec, knots=knots, degree=mspline_degree,
                                intercept=FALSE, Boundary.knots = c(0, time_max))
log_ratios2 <- -post2[,,sprintf("log_lik[%d]",1:N)]
r_eff2 <- relative_eff(exp(log_ratios2))
psis_result2 <- psis(log_ratios2, r_eff = r_eff2)
w2 <- weights(psis_result2, log=FALSE) #note that colSums(w2) = rep(1, dim(w2)[2])
survs_loo2 <- surv_mspline_loo(i_spline_basis_evals,
                            X,
                            matrix(post2[,,sprintf("gammas[%d]", 1:nbasis)], dim(post2)[1]*dim(post2)[2], nbasis),
                            as.vector(post2[,,"intercept"]),
                            matrix(post2[,,sprintf("betas[%d]" ,1:ncol(X))], dim(post2)[1]*dim(post2)[2], ncol(X)),
                            t(w2)
                            )

survs2 <- surv_mspline_loo(i_spline_basis_evals,
                            X,
                            matrix(post2[,,sprintf("gammas[%d]", 1:nbasis)], dim(post2)[1]*dim(post2)[2], nbasis),
                            as.vector(post2[,,"intercept"]),
                            matrix(post2[,,sprintf("betas[%d]" ,1:ncol(X))], dim(post2)[1]*dim(post2)[2], ncol(X)),
                            matrix(rep(1, dim(w2)[1]*dim(w2)[2]), nrow = dim(w2)[2], byrow = T)/dim(w2)[1]
                            )
```

```{r, echo=FALSE}
pec_rslt <- pec(list("MSpline-Loo"=survs_loo2, 
                     "Constant-Loo"=survs_loo1,
                     "MSpline"=survs2,
                     "Constant"=survs1
                     ), formula =Surv(time,event)~metastized , data=df, exact = F, times = times_pec,cens.model="marginal",
                splitMethod="none",
                B=0,
                verbose=TRUE)

df_pec <- tibble(time=pec_rslt$time, bs=pec_rslt$AppErr[["Constant-Loo"]], model="Constant-Loo") %>%
  bind_rows(tibble(time=pec_rslt$time, bs=pec_rslt$AppErr[["MSpline-Loo"]], model="MSpline-Loo")) %>%
  bind_rows(tibble(time=pec_rslt$time, bs=pec_rslt$AppErr[["Constant"]], model="Constant")) %>%
  bind_rows(tibble(time=pec_rslt$time, bs=pec_rslt$AppErr[["MSpline"]], model="MSpline")) %>%

  bind_rows(tibble(time=pec_rslt$time, bs=pec_rslt$AppErr[["Reference"]], model="Kaplan-Meier"))
ggplot(data=df_pec) +
  geom_hline(yintercept=.25,alpha=.5, color='gray', linetype='dashed')+
  geom_vline(xintercept = c(0,knots, time_max), alpha=.5, color='gray', linetype='dashed')+
  geom_step(aes(x=time, y=bs, color=model))+
  geom_rug(data=tibble(times=times), aes(x=times), col="steelblue",alpha=0.5, size=1.25)+
  ylab("Brier Score")+
  xlab("Time")+  
  scale_fill_manual(values=cbbPalette)+
  scale_colour_manual(values=cbbPalette)

trsf_bs <- function(m) {
  dplyr::filter(df_pec, model==m) %>%
    dplyr::select(time, bs) %>%
    dplyr::rename(!!paste("bs",m,sep="_") := bs)
}

knitr::kable(
purrr::map(unique(df_pec$model),trsf_bs) %>% 
  purrr::reduce(left_join, by = "time")
)

```

## Constant only 

```{r, echo=FALSE}
pec_rslt <- pec(list("Constant-Loo"=survs_loo1, 
                     "Constant"=survs1
                     ), formula =Surv(time,event)~metastized , data=df, exact = F, times = times_pec,cens.model="marginal",
                splitMethod="none",
                B=0,
                verbose=TRUE)

df_pec <- tibble(time=pec_rslt$time, bs=pec_rslt$AppErr[["Constant-Loo"]], model="Constant-Loo") %>%
  bind_rows(tibble(time=pec_rslt$time, bs=pec_rslt$AppErr[["Constant"]], model="Constant"))

ggplot(data=df_pec) +
  geom_hline(yintercept=.25,alpha=.5, color='gray', linetype='dashed')+
  geom_step(aes(x=time, y=bs, color=model))+
  geom_rug(data=tibble(times=times), aes(x=times), col="steelblue",alpha=0.5, size=1.25)+
  ylab("Brier Score")+
  xlab("Time")+  
  scale_fill_manual(values=cbbPalette)+
  scale_colour_manual(values=cbbPalette)
knitr::kable(
purrr::map(unique(df_pec$model),trsf_bs) %>% 
  purrr::reduce(left_join, by = "time")
)
```

## MSpline only 

```{r, echo=FALSE}
pec_rslt <- pec(list("MSpline-Loo"=survs_loo2, 
                     "MSpline"=survs2
                     ), formula =Surv(time,event)~metastized , data=df, exact = F, times = times_pec,cens.model="marginal",
                splitMethod="none",
                B=0,
                verbose=TRUE)

df_pec <- tibble(time=pec_rslt$time, bs=pec_rslt$AppErr[["MSpline-Loo"]], model="MSpline-Loo") %>%
  bind_rows(tibble(time=pec_rslt$time, bs=pec_rslt$AppErr[["MSpline"]], model="MSpline"))

ggplot(data=df_pec) +
  geom_hline(yintercept=.25,alpha=.5, color='gray', linetype='dashed')+
  geom_vline(xintercept = c(0,knots, time_max), alpha=.5, color='gray', linetype='dashed')+
  geom_step(aes(x=time, y=bs, color=model))+
  geom_rug(data=tibble(times=times), aes(x=times), col="steelblue",alpha=0.5, size=1.25)+
  ylab("Brier Score")+
  xlab("Time")+  
  scale_fill_manual(values=cbbPalette)+
  scale_colour_manual(values=cbbPalette)
knitr::kable(
purrr::map(unique(df_pec$model),trsf_bs) %>% 
  purrr::reduce(left_join, by = "time")
)
```

